---
title: "info201 PS05"
author: "Chenyang Yuan"
date: "2023-02-19"
output: html_document
---

```{r setup, include=FALSE}
library(tidyverse)
knitr::opts_chunk$set(echo = TRUE)
```

## Load and check date

```{r}
gapminder <- read_delim("data/gapminder.csv")
```  
  
**How many rows/columns do we have?**
```{r}
ncolumn <- ncol(gapminder)
nrow <- nrow(gapminder)
```   

There are `r ncolumn` columns and `r nrow` rows of data.


**Print a small sample of data. Does it look OK?**
```{r}
gapminder %>%
  sample_n(5)
```

The data looks good. Each column labeled correctly.

## Descriptive statistics
**How many countries are there in the dataset? Analyze all three: iso3, iso2 and name.**
```{r}
gapminder %>%
  summarize(num_iso2 = n_distinct(iso2),
            num_iso3 = n_distinct(iso3),
            num_name = n_distinct(name))
```
There are _249_ 2-letter country codes, _253_ 3-letter country codes, 
and _250_ unique country names.

**Find how many names are there for each iso-2 code. Are there any iso-2 codes that correspond to more than one name? What are these countries?**
```{R}
gapminder %>%
  group_by(iso2) %>%
  summarize(num = n_distinct(name)) %>%
  arrange(desc(num)) %>%
  head(5)
```
2-letter code **NA** corresponds to the names of two countries.

```{R}
gapminder %>%
  group_by(iso2) %>%
  filter(is.na(iso2)) %>%
  summarize(name = unique(name))
```
The countries whose 2-letter code is NA are Namibia and NA, but I think the second country NA means Null Value.

**Now repeat the same for name and iso3-code. Are there country names that have more than one iso3-code? What are these countries?**
```{R}
gapminder %>%
  group_by(name) %>%
  summarize(num = n_distinct(iso3)) %>%
  arrange(desc(num)) %>%
  head(5)
```
The country name **NA** corresponds to four 3-letter country codes.

```{R}
gapminder %>%
  group_by(name) %>%
  filter(is.na(name)) %>%
  summarize(three_letter = unique(iso3))
```
The countries whose name is NA but have 3-letter code are CHANISL, GBM, KOS, NLD_CURACAO.

**What is the minimum and maximum year in these data?**
```{r}
max <- max(gapminder$time, na.rm = TRUE)
min <- min(gapminder$time, na.rm = TRUE)
```
The minimum year is `r min`, and the maximum year is `r max`

## CO2 emissions
**How many missing co2 emissions are there for each year? Analyze both missing CO2 and co2_PC. Which years have most missing data?**
```{r}
gapminder %>% 
  group_by(time) %>% 
  filter(is.na(co2) | is.na(co2_PC)) %>%
  summarize(num_miss = length(time)) %>%
  arrange(desc(num_miss)) %>%
  head(5)
```
The years _2017, 2018, and 2019_ both have the most amount of missing data with _217_.

**Make a plot of total CO2 emissions over time for the U.S, China, and India. Add a few more countries of your choice. Explain what do you see.**
```{r}
gapminder %>%
  filter(name == "United States of America" | 
         name == "China" | 
         name == "India" |
         name == "France" | 
         name == "Japan",
         !is.na(co2)) %>%
  ggplot() + 
  geom_line(aes(x = time, y = co2, col = name))
```
<br> China's CO2 emission **increased the most** from 1960 to 2020. During this period, France's CO2 emission were the most stable, with no large increase or decrease.

**Now let’s analyze the CO2 emissions per capita (co2_PC ). Make a similar plot of the same countries. What does this figure suggest?**
```{r}
gapminder %>%
  filter(name == "United States of America" |
         name == "China" | 
         name == "India" | 
         name == "France" | 
         name == "Japan",
         !is.na(co2_PC)) %>%
  ggplot() + 
  geom_line(aes(x = time, y = co2_PC, col = name))
```
<br> United States has the **highest carbon emission per capita**, and India has the **lowest carbon emission per capita**.

**Compute average CO2 emissions per capita across the continents (assume region is the same as continent). Comment what do you see.**
```{r}
gapminder %>% 
  filter(!is.na(co2_PC), 
         !is.na(region)) %>%
  group_by(region, time) %>% 
  summarize(average_co2_PC = mean(co2_PC, na.rm = TRUE))  %>%
  arrange((average_co2_PC)) %>%
  head(5)
```
Africa continent had the **lowest average CO2 emission per capita** in 1960, and the lowest emissions from 1960-1964 were in Africa.

**Make a barplot where you show the previous results–average CO2 emissions per capita across continents in 1960 and 2016.**
```{r, message= FALSE}
gapminder %>% 
  filter(time == 1960 | time == 2016, 
         !is.na(co2_PC), 
         !is.na(region) ) %>%
  filter(region != "") %>% 
  group_by(time, region) %>% 
  summarize(average_co2_PC = mean(co2_PC, na.rm = TRUE)) %>%

  ggplot() + 
  geom_bar(aes(x = region,
               y = average_co2_PC, 
               fill = factor(time)),
           stat = "identity",
           position = "dodge")
```
<br> This graph shows that CO2 emissions in Asia **increased the most** from _1960 to 2016_, while the Americas **decreased** during this period.

**Which countries are the three largest, and three smallest CO2 emitters (in terms of CO2 per capita) in 2016 for each continent?**

```{r}
gapminder %>%
  filter(time == 2016, 
         !is.na(co2_PC), 
         !is.na(region),
         !is.na(name)) %>%
  group_by(region, name) %>%
  summarize(co2_pc_data = co2_PC) %>%
  arrange(region, desc(co2_pc_data)) %>%
  mutate(row = row_number()) %>%
  filter(row_number() %in% c(1,2,3,n(),n()-1,n()-2)) %>%
  ungroup() %>%
  arrange(region, row)
```
**From largest to smallest**:

Africa: South Africa, Libya, Seychelles, Burundi, Somalia, Congo: Democratic Republic of the

Americas: Trinidad and Tobago, United States of America, Canada, Honduras, Nicaragua, Haiti

Asia: Qatar, Kuwait, Bahrain, Yemen, Nepal, Afghanistan

Europe: Gibraltar, Luxembourg, Faroe Island, Moldova: Republic of, Albania, Liechtenstein 

Oceania: New Caledonia, Australia, Palau, Kiribati, Vanuatu, Solomon Islands 

## GDP per capita
**Make a scatterplot of GDP per capita versus life expectancy by country, using data for 1960. Make the point size dependent on the country size, and color those according to the continent. Feel free to adjust the plot in other ways to make it better.**
```{r}
gapminder %>% 
  filter(time == 1960,
         !is.na(GDP_PC), 
         !is.na(lifeExpectancy), 
         !is.na(name)) %>%
  ggplot() +
  geom_point(aes(x = GDP_PC, 
                 y = lifeExpectancy, 
                 col = region,
                 size = totalPopulation))
```
<br> In 1960, Europe has highest life expectancy and Africa has lowest life expectancy. _The hight the GDP per capita, the higher the life expectancy_.

**Make a similar plot, but this time use 2019 data only.**
```{r}
gapminder %>% 
  filter(time == 2019,
         !is.na(GDP_PC), 
         !is.na(lifeExpectancy), 
         !is.na(name)) %>%
  ggplot() +
  geom_point(aes(x = GDP_PC, 
                 y = lifeExpectancy, 
                 col = region,
                 size = totalPopulation))
```
<br> In 2019, Europe still has highest life expectancy.

**Compare these two plots and comment what do you see. How has world developed through the last 60 years?**   
Life expectancy in all continents has increased. Asia sees **biggest increase** in life expectancy.

**Compute the average life expectancy for each continent in 1960 and 2019. Do the results fit with what do you see on the figures?**

```{r}
gapminder %>% 
  filter(time == 1960 | time == 2019,
         !is.na(lifeExpectancy),
         !is.na(name)) %>% 
  group_by(time, region) %>% 
  summarize(average_le = mean(lifeExpectancy, na.rm = TRUE)) %>% 
  arrange(time, desc(average_le))
```
Yes. Europe has highest average LE, and Africa has lowest average LE in 1960 and 2019.

**Compute the average LE growth from 1960-2019 across the continents. Show the results in the order of growth. Explain what do you see.**
```{r}
gapminder %>%
  filter(time == 1960 | time == 2019,
         !is.na(region),
         !is.na(name),
         !is.na(lifeExpectancy)) %>%
  group_by(region, time) %>%
  summarize(average_le = mean(lifeExpectancy)) %>%
  mutate(difference = average_le - lag(average_le)) %>%
  group_by(region) %>%
  summarize(average_le_growth = mean(difference, na.rm = TRUE)) %>%
  arrange(average_le_growth)
```
Europe has **lowest LE growth**, and Asia has **highest LE growth**.

**Show the histogram of GDP per capita for years of 1960 and 2019. Try to put both histograms on the same graph, see how well you can do it!**
```{r}
gapminder %>% 
  filter(time == 1960 | time == 2019, 
         !is.na(GDP_PC)) %>% 
  ggplot() + 
  geom_histogram(aes(x = GDP_PC, 
                     fill = factor(time)),
                 position = "dodge",
                 bins = 5)
```
<br>GDP per capita has **increased** in the 60 years.

**What was the ranking of US in terms of life expectancy in 1960 and in 2019? (When counting from top.)**
```{r}
gapminder %>%
  filter(time == 1960 | time == 2019,
         !is.na(name),
         !is.na(lifeExpectancy)) %>%
  group_by(time) %>%
  mutate(le_rank = rank(desc(lifeExpectancy))) %>%
  filter(name == "United States of America") %>%
  summarize(name, le_rank)
```
The ranking of US in terms of LE is 17 in 1960 and 46 in 2019

**If you did this correctly, then you noticed that US ranking has been falling quite a bit. But we also have more countries in 2019–what about the relative rank divided by the corresponding number of countries that have LE data in the corresponding year?**
```{r}
gapminder %>%
  filter(time == 1960 | time == 2019,
         !is.na(name),
         !is.na(lifeExpectancy)) %>%
  group_by(time) %>%
  mutate(le_rank = rank(desc(lifeExpectancy))) %>%
  mutate(ratio = le_rank / n()) %>%
  filter(name == "United States of America") %>%
  summarize(name, le_rank, ratio)
```
The ratio of the US in 2019 is still not as high as it was in 1960